Autonomous Oceanographic Data Collection towards Estimating Excursion Sets of Vector-valued Random Fields

### 2020-06-09 main_new.tex
Updates from Overleaf

% "Title of the paper"
\title{Autonomous Oceanographic Sampling Designs using Excursion Sets for Multivariate Gaussian random fields}
\runtitle{Excursion Probabilities for Informative Sampling}

\begin{aug}
\author{\fnms{Trygve Olav} \snm{Fossum}\thanksref{t1,t2}, \corref{} \ead[label=e1]{trygve.o.fossum@ntnu.no}}
\author{\fnms{Cédric} \snm{Travelletti}\thanksref{t3}, \corref{} \ead[label=e2]{cedric.travelletti@stat.unibe.ch}}
\author{\fnms{Jo} \snm{Eidsvik}\thanksref{t4}, \ead[label=e3]{jo.eidsvik@ntnu.no}}
\author{\fnms{David} \snm{Ginsbourger}\thanksref{t3}, \ead[label=e4]{david.ginsbourger@idiap.ch}}
\and
\author{\fnms{Kanna} \snm{Rajan}\thanksref{t5}. \ead[label=e5]{kanna.rajan@fe.up.pt}}

\affiliation[t1]{Department of Marine Technology, The Norwegian University of Science and Technology (NTNU), Trondheim, Norway.} 
\affiliation[t2]{Centre for Autonomous Marine Operations and Systems (AMOS).}
\affiliation[t3]{Institute of Mathematical Statistics and Actuarial Science, University of Bern, Switzerland.}
\affiliation[t4]{Department of Mathematical Sciences, NTNU.}
\affiliation[t5]{Underwater Systems and Technology Laboratory, Faculty of Engineering, University of Porto, Portugal.}

@@

\end{frontmatter}

\section{Introduction}

@@

The remainder of this paper is organized as follows: Section
\ref{sec:bg} provides select background on ocean sampling. Section
\ref{sec:ESEP} defines ESs, EPs, and the design criteria of IBV for
vector-valued GPs. Section \ref{sec:heuristics} builds on these
assumptions when deriving the sequential design criteria for adaptive
sampling. Section \ref{sec:simulations} discusses properties of the
methods in simulation studies. Section \ref{sec:case_study}
demonstrates the methodology used in field work characterizing a river
plume and finally, Section \ref{sec:concl_disc} contains a summary and
a discussion of future work.

\section{Ocean Processes and Observation}
\label{sec:bg}

@@

The AUVs considered here are powered untethered platforms, that
operate at $1$-$3$ m/s in the upper water column, where they are free
to move in six degrees of freedom (6 DOF). The in-water operation time
capability depends on survey speed, payload sensors and navigation;
typically this amounts to 8-48 hrs. AUVs typically use single-board
computers (SBCs), like the Raspberry Pi or a multicore GPU NVIDIA
Jetson TX1 (quad-core 1.91 GHz 64-bit ARM machine, a 2-MB L2 shared
cache, and 4 GB of 1600 MHz DRAM) for computation onboard.

Surveys with AUVs are usually limited to observations along fixed
transects that are pre-scripted in mission plans created manually by a
human operator. The mission can be specified operating on a scale of
hundreds of meters to tens of kilometers depending on the scientific
context. Faced with limited coverage capacity, a more effective
approach is to instead use onboard algorithms to continuously
evaluate, update, and refine future sampling locations (sense-plan-act
cycle in Fig. \ref{fig:envir1}), making the information gathering
\emph{adaptive} \citep{das11b,Das2015,fossuminformation,fossum18b}.
% This usually occurs on a spatial grid, called a waypoint graph. 
In doing so, the space of sampling opportunities is still limited by
the waypoint graph, but the AUV has flexibility to modify its path at
each waypoint based on in-situ sensing and measurements onboard
\citep{py10,Rajan12,Rajan12b}.

@@

Assuming now that $q$ observations are available, one wishes to
predict $\gp[\x]$. The multivariate Kriging equations holds with size
$p$ vector of conditional means:  
\begin{equation}\label{eq:cokrig}
\bbm(\x)=\bmu(\x)+\lambda(\x,\bu_d)^T \left( \by_d-G_d \bmu(\bu_d)\right),
\end{equation}
where $\bmu$ is size $p$ unconditional mean function of $Z$, $\by_d$
is the realized vector of observed responses, and
$\lambda(\x,\bu_d)=K^{-1}_y k(\bu_d, \bu)$. Here,
the data covariance is
$K_y=\mbox{Cov}(\bY_d,\bY_d)=G_d K(\bu_d,\bu_d)G_d^T+R_d$ while
the cross-covariance is $\mbox{Cov}(\bY_d,Z(\x))=G_d k(\bu_d, \x)$, assuming the measurement
error is independent of the random fields.  The Kriging (cross-)covariance function can also be expressed in the same vein via
a $p \times p$ matrix 
\begin{equation}\label{eq:krig:cov}
\Sigma(\x,\x')=K(\x,\x')-\lambda(\x,\bu_d)^T K_y(\bu_d, \bu_d) \lambda(\x,\bu_d).
\end{equation}

@@

A critical element in the derivation of a closed-form for the EIBV is
that the conditional mean in Eq. \eqref{eq:cokrig} is a linear
(affine) function of the data $\by_d$ and the covariance is not a
function of the data. This in turn means that the probabilities in Eq.
\eqref{eq:post_ep} involve inequality statements for linear
combinations of Gaussian variables. Related closed-form solutions have
been noted in similar contexts \citep{bhattacharjya2013value,
  chevalier2014fast,stroh}, but not generalized to our situation with
random sets for vector-valued GPs. Notably, the mean in equation
\eqref{eq:cokrig}, is in the form of $\ba+B V$, where
$V \sim N(0_q, K_y (\bu_d,\bu_d))$,
$\ba=(t_1,t_2)^t-\bmu(\x)+\lambda(\x,\bu_d)^T G_d \bmu(\bu_d)$
and $B=\lambda(\x,\bu_d)^T$. The following then applies to give a
closed-form solution to EIBV calculation in this case.

@@

The design and data define
the conditional GP with mean
$E(\gp[\x]|\mathcal{Y}^{s-1})$ and (cross) covariance matrix
$\mbox{Var}(\gp[\x]|\mathcal{Y}^{s-1})$. These are efficiently updated in a sequential manner keeping track of the mean and covariance kernel of the GP over the stages. As extensions of Eq. \eqref{eq:cokrig} and \eqref{eq:krig:cov}, we have
\begin{eqnarray}\label{eq:krigSUR}
\mu^{s}(\x) &=&\mu^{s-1}(\x)+\lambda^{[s,s]}(\x,\bu_{d^{j,s}})^T (\by_{d^{j,s}}-G_{d^{j,s}} \mu(\x)), \nonumber \\
k^{s}(\x,\x') &=& k^{s-1}(\x,\x')-\lambda^{[s,s]}(\x,\bu_{d^{j,s}})^T K_{Y_{d^{j,s}}} \lambda^{[s,s]}(\x',\bu_{d^{j,s}}),
\end{eqnarray}
where $\lambda^{[s,s]}(\x,\bu_{d^{j,s}})=K^{-1}_{Y_{d^{j,s}}} k(\bu_{d^{j,s}},\x)$ and the data covariance is $K^{-1}_{Y_{d^{j,s}}}=G_{d^{j,s}} K^s(\bu_{d^{j,s}},\bu_{d^{j,s}})G^T_{d^{j,s}}+R_{d^{j,s}}$, with measurement model and design matrix defined as in equation \eqref{eq:meas}

At stage $s$, the robot must then choose design $d^{j,s}$ where measurements $\by_{d^{j,s}}$ should
be gathered. For the triangularized waypoint
graph all six possible
directions must be evaluated to minimize the expected uncertainty of random sets, and in particular the EIBV for our 
application. We hence focus on choosing designs for data modeled by
equation \eqref{eq:meas}, to conduct efficient sequential uncertainty
reduction according to the EIBV criterion.  Notably, compared with the
static design $d$ in equation \eqref{eq:eibv}, the optimal sequential
design not only considers the best current design grid point, but also
how the data gathered at this node could inform future sampling at
successive stages.

@@

Even though the strategy looks at two sampling stages, it is only used to find
the current best design. When data are collected at stage $s$, the GP model is
updated as in Eqn. \eqref{eq:krigSUR}, and the mean and (cross) covariance are
used to compute the next design for stage $s+1$, now anticipating what
stage $s+1$ and $s+2$ could offer, and so on.
This look-ahead approach is much more computationally demanding than
the myopic strategy, so for practical implementation onboard the AUV this is reaching the limits allowed for calculation. We conduct some pruning of the least valuable design paths in the evaluation of Eq. \eqref{critLA}, to get a satisfactory computation time below one minute per waypoint. This means that we use the myopic
strategy to rank the three best designs following the first stage alone, and
for each of these preferred designs, we undertake the look-ahead calculations.

>>>>

% "New Title of the paper"
\title{Autonomous Oceanographic Data Collection towards Estimating Excursion Sets of Vector-valued Random Fields} 
\runtitle{Autonomous Oceanographic Data Collection}

% "Former Title of the paper" (DG 08.06.2020)
%\title{Autonomous Oceanographic Sampling Designs using Excursion Sets for Multivariate Gaussian random fields} \runtitle{Excursion Probabilities for Informative Sampling}

\begin{aug}
\author{\fnms{Trygve Olav} \snm{Fossum}\thanksref{t1,t2}, \corref{} \ead[label=e1]{trygve.o.fossum@ntnu.no}}
\author{\fnms{Cédric} \snm{Travelletti}\thanksref{t3}, \corref{} \ead[label=e2]{cedric.travelletti@stat.unibe.ch}}
\author{\fnms{Jo} \snm{Eidsvik}\thanksref{t4}, \ead[label=e3]{jo.eidsvik@ntnu.no}}
\author{\fnms{David} \snm{Ginsbourger}\thanksref{t3}, \ead[label=e4]{david.ginsbourger@stat.unibe.ch}}
\and
\author{\fnms{Kanna} \snm{Rajan}\thanksref{t5}. \ead[label=e5]{kanna.rajan@fe.up.pt}}

\affiliation[t1]{Department of Marine Technology, The Norwegian University of Science and Technology (NTNU), Trondheim, Norway.} 
\affiliation[t2]{Centre for Autonomous Marine Operations and Systems, NTNU.}
\affiliation[t3]{Institute of Mathematical Statistics and Actuarial Science, University of Bern, Switzerland.}
\affiliation[t4]{Department of Mathematical Sciences, NTNU.}
\affiliation[t5]{Underwater Systems and Technology Laboratory, Faculty of Engineering, University of Porto, Portugal.}

@@

\end{frontmatter}
\section{Introduction}

%NB: The subsections are used as a temporary instrument to highlight the proposed structure and should probably be abolished upon convergence to a stable version.

\subsection{Oceanic data collection and spatial design of experiments}

%\textcolor{blue}{NB: I started to match previously listed thoughts/ideas with pieces of the former intro. Will do more asap. Also, Section 3 I find excellent (and more precise than the intro) and start thinking that some bits might end up in the newer intro (while some others could land in a later section).}

Monitoring the world's oceans has gained increased importance in light of the changing climate and increasing anthropogenic impact. 
\begin{comment} % TO BE RE-INSERTED LATER IN THE ETXT?
A central problem for understanding these factors is the lack of representative data with sufficient resolution. Most of this \emph{undersampling} can be attributed to the large spatio-temporal variations in which ocean processes transpire, prompting the need for effective means of sampling. By \emph{sampling}, we refer to the design of observational strategies in the spatial domain, where the use autonomous robotic platforms can be combined with statistical methods to pursue measurements with high scientific relevance. This combination, of statistical tools and robotic platforms, constitute a methodological basis for addressing the problem of efficient sampling of the ocean, which is at the core of this work.
\end{comment}
Central to understanding the changes taking place in the upper
water-column is knowledge of the bio-geophysical interaction driven by
an agglomeration of physical forcings (e.g. wind, topography,
bathymetry, tidal influences, etc.) and incipient micro-biology driven
by planktonic and coastal anthropogenic input, such as pollution and
agricultural runoff transported into the ocean by rivers and streams.
These often result in a range of ecosystem-related phenomena such as
blooms and plumes, with direct and indirect effects on society. \kc{we
  should look for a citation here. Check with John Ryan}
 One of the bottlenecks in the study of such phenomena lies however in the lack of obervational data with sufficient resolution. Most of this \emph{undersampling} can be attributed to the large spatio-temporal variations in which ocean processes transpire, prompting the need for effective means of data collection. Based on collected data and sometimes also of prior knowledge, models from spatial statistics can be used to make predictions of the state variables under consideration over regions of interest. In case of multiple continuous state variables being jointly studied relying on spatially scattered data, vector-valued random field models and associated prediction approaches such as co-Kriging (REFs) can be considered nowadays as part of the standard statistical toolkit in oceanographic research and beyond. %applications and beyond.  
 %By \emph{sampling}, we refer to the design of observational strategies in the spatial domain, where the use autonomous robotic platforms can be combined with statistical methods to pursue measurements with high scientific relevance. This combination, of statistical tools and robotic platforms, constitute a methodological basis for addressing the problem of efficient sampling of the ocean, which is at the core of this work.

Traditional data collection to understand natural and artificial phenomena at sea has typically been based on static buoys, floats, or ship-based methods,
with significant logistical limitations that directly impact coverage
and sampling resolution. Modern methods using satellite remote-sensing
provide large-scale coverage but have limited resolution, are limited
to sensing the surface, and are impacted by cloud cover. Numerical
ocean models similarly find it challenging to provide detail at fine
scale \citep{Lermusiaux:2006}, and also come with computation costs that can be practically limiting. 
%limiting computing resources, no time for hard-core ocean model Navier-Stokes, hence GPs and approximations, etc.
%
\begin{figure}[!h] 
  \centering 
  \subfigure[Illustration of an ocean sensing network and the sense-plan-act control methodology.]{\includegraphics[width =
    0.49\textwidth]{Figures/envir2.pdf}\label{fig:envir1}}
  \hfill
  \subfigure[Frontal patterns off of the Nidelva river, Trondheim, Norway.]{\includegraphics[width =
    0.49\textwidth]{Figures/river_proccess.pdf}\label{fig:nidelven}}
  \caption{\ref{fig:envir1} Traditional ocean observation based on 
    ship-based sampling has been augmented by autonomous
    robotic vehicles. % and their interactions.  
    AUV platforms are an integral part of this network being able to
    reason and make decisions for efficient onboard adaptive sampling.
    % using the sense-plan-act control approach to autonomous control. 
    \ref{fig:nidelven} The interaction of river and ocean creates
    processes that are challenging to map, where the combination of
    statistics and robotics can play a vital role in enabling more
    effective oceanographic observation.\kc{The inset ``AUV robot''
      looks very tacky and ad-hoc; might be best to remove it. Change
      ``ships'' to
      ``Ship'', remove the aerial drone, since we didn't use it and
      don't plan to in this work.}}
  \label{fig:envir} \end{figure}
%
\begin{comment} % Somehow redundant with the parag following below
Mobile robotic platforms \citep{Bellingham07} have contributed significantly to environmental monitoring and sampling. In particular, AUVs have advanced the state of in-situ sampling, and made robotics an integral part of ocean observation, filling parts of the undersampling gap \cite{rudnick03,rudnick18} (See Fig. \ref{fig:envir1}).
\end{comment}
%
The advent of robust mobile robotic platforms \citep{Bellingham07} has
resulted in significant contributions to environmental monitoring and
sampling in the ocean. In particular, autonomous underwater vehicles (AUVs) have advanced the state of data collection and consequently have made robotics an integral part of ocean observation; %our 
previous work by some of the authors of the present paper have contributed to this effort \citep{das11b,Das2015,fossuminformation,fossum18b}.

As full numerical ocean models based on complex differential equations cannot be run onboard a marine robot with limited computational capacity, statistical models relying on collected data and random field assumptions appear quite relevant as means to guide AUV data collection trajectories. Our main research angle in the present work is to extend sequential design strategies from the world of spatial statistics and computer experiments to frameworks featuring both vector-valued observational data and experimental designs that may be constrained by the nature of measurement paths.      
\\

Surveys with AUVs are usually limited to observations along fixed
transects that are pre-scripted in mission plans created manually by a
human operator. Missions can be specified operating on a scale of
hundreds of meters to tens of kilometers depending on the scientific
context. Faced with limited coverage capacity, a more effective
approach is to instead use onboard algorithms to continuously
evaluate, update, and refine future sampling locations (sense-plan-act
cycle in Fig. \ref{fig:envir1}), making the information gathering
\emph{adaptive} \citep{das11b,Das2015,fossuminformation,fossum18b}.
% This usually occurs on a spatial grid, called a waypoint graph. 
In doing so, the space of sampling opportunities is still limited by
the so-called waypoint graph, a discretization of the search domain, but the AUV has flexibility to modify its path at each waypoint based on in-situ sensing and measurements onboard \citep{py10,Rajan12,Rajan12b}.\\

The work presented here is primarily inspired by a cast study 
%application where an AUV is used to measure both temperature and salinity in 
pertaining to the spatial characterization of a frontal system generated by
river plumes. Fig. \ref{fig:nidelven} shows the survey area in
Trondheim, Norway, where cold freshwater enters from the river,
creating a strong gradient in both temperature and salinity. Because
of the local topography and the Coriolis force the cold fresh water
tends to flow to the east. Depending on the river discharge, tidal
effects, wind, and temperature differences, this boundary often gets
distorted. Initial knowledge about the location and evolution of these
features are highly uncertain, making deterministic planning
challenging. Here the goal is to use AUV measurements towards a better identification of freshwater versus oceanic water, with the help of a vector-valued random field model for both temperature and salinity. As we will present in the next sections, some of the original contributions inspired by this specific problem are actually applicable to further settings and application areas. \\

The questions tackled here hence pertain to the broader area of spatial data collection and experimental design, yet with the particularity of estimating some regions of the domain, mostly excursion sets, being implicitly defined by a scarcely observed vector-valued random field. 
%because of limited resources to sample the large oceanographic domains, one must plan for active learning during the operation, where a relevant criterion is used to extract the most valuable designs. The acquired in-situ information must then be assimilated into the statistical model onboard the platform, so that it can be used to re-compute the sampling criterion and in this way inform decisions on where to sample next. \\
Moreover, given constraints on AUV movements and the fact that surveya rely by design in such a framework on successive measurements spread along a trajectory, addressing corresponding design problems calls for particular sequential strategies, possibly also accounting for the finiteness of resources by anticipating multiple steps ahead thereby improving over myopic strategies. \\

Here we aim to leverage and extent recent progress in uncertainty quantification and reduction for excursion sets of Gaussian random fields in order to address vector-valued observational cases under constrained design such as in the motivating application of designing AUV trajectories to better distinguish between freshwater from ocean waters by relying on temperature and salinity measurements. %approach will take substantial advantage of existing work in the field of excursion set estimation mainly dedicated to computer experiments in the scalar-valued case, and where space exploration can be performed without constraints regarding the distance between successive design points. 
Original contributions compared to recent related work in the framework of scalar-valued computer experiments include: 
\begin{itemize}
    \item Extension of two important Sequential Uncertainty Reduction criteria of (REFs) to homotopic or heterotopic vector-valued settings,
    \item Investigation of myopic and multiple-step ahead batch-sequential strategies for optimizing trajectories with respect to such criteria,
    \item Replicable experiments with synthetic test cases and original data sets with accompagnying codes 
\end{itemize}
Let us briefly review random field modelling and recent advances in targetd sequential design of experiments based on Gaussian Processes before detailing further existing literature tackling the considered problems, our proposed approach, as well as outlining the rest of the paper.

\subsection{Random field modelling and targeted sequential design of experiments}

\begin{itemize}
    \item 
\end{itemize}

% From earlier version; to be relocated?
%Hence statistical proxy models of the environment must be used. For our purpose, we rely on Gaussian process (GP) representations of the ocean variables of interest because they are computationally convenient and yet provide enough flexibility to realistically model the spatial variability and dependence, and the correlation between multiple processes. 
\subsection{Related work and proposed approach in a nutshell}

\begin{itemize}
    \item 
\end{itemize}

\textcolor{blue}{We might want to re-use some of the following (from earlier version):}

\subsection{Outline of the paper and its appendices / supplemental material}

\begin{itemize}
    \item 
\end{itemize}

The remainder of this paper is organized as follows: 
%Section \ref{sec:bg} provides select background on ocean sampling. 
Section \ref{sec:ESEP} defines ESs, EPs, and the design criteria of IBV for
vector-valued GPs. Section \ref{sec:heuristics} builds on these
assumptions when deriving the sequential design criteria for adaptive
sampling. Section \ref{sec:simulations} discusses properties of the
methods in simulation studies. Section \ref{sec:case_study}
demonstrates the methodology used in field work characterizing a river
plume and finally, Section \ref{sec:concl_disc} contains a summary and
a discussion of future work.

%\subsection{}

%\begin{itemize}
%    \item 
%\end{itemize}

\begin{comment}
\section{Former Introduction (Before 08.06.2020)}

@@

The remainder of this paper is organized as follows: Section
\ref{sec:bg} provides select background on ocean sampling. Section
\ref{sec:ESEP} defines ESs, EPs, and the design criteria of IBV for
vector-valued GPs. Section \ref{sec:heuristics} builds on these
assumptions when deriving the sequential design criteria for adaptive
sampling. Section \ref{sec:simulations} discusses properties of the
methods in simulation studies. Section \ref{sec:case_study}
demonstrates the methodology used in field work characterizing a river
plume and finally, Section \ref{sec:concl_disc} contains a summary and
a discussion of future work.
\end{comment}

\begin{comment}
\section{Ocean Processes and Observation}
\label{sec:bg}

@@

\subsection{To be relocated elsewhere}
The AUVs considered here are powered untethered platforms, that
operate at $1$-$3$ m/s in the upper water column, where they are free
to move in six degrees of freedom (6 DOF). The in-water operation time
capability depends on survey speed, payload sensors and navigation;
typically this amounts to 8-48 hrs. AUVs typically use single-board
computers (SBCs), like the Raspberry Pi or a multicore GPU NVIDIA
Jetson TX1 (quad-core 1.91 GHz 64-bit ARM machine, a 2-MB L2 shared
cache, and 4 GB of 1600 MHz DRAM) for computation onboard.

\begin{comment}
Surveys with AUVs are usually limited to observations along fixed
transects that are pre-scripted in mission plans created manually by a
human operator. The mission can be specified operating on a scale of
hundreds of meters to tens of kilometers depending on the scientific
context. Faced with limited coverage capacity, a more effective
approach is to instead use onboard algorithms to continuously
evaluate, update, and refine future sampling locations (sense-plan-act
cycle in Fig. \ref{fig:envir1}), making the information gathering
\emph{adaptive} \citep{das11b,Das2015,fossuminformation,fossum18b}.
% This usually occurs on a spatial grid, called a waypoint graph. 
In doing so, the space of sampling opportunities is still limited by
the waypoint graph, but the AUV has flexibility to modify its path at
each waypoint based on in-situ sensing and measurements onboard
\citep{py10,Rajan12,Rajan12b}.

@@

Assuming now that $q$ observations $\by_d$ made at $\bu_d$ are available, one wishes to predict $\gp[\x]$. The multivariate Kriging equations holds with size
$p$ vector of conditional means:  
\begin{equation}\label{eq:cokrig}
\bbm(\x)=\bmu(\x)+\lambda(\x,\bu_d)^T \left( \by_d-G_d \bmu(\bu_d)\right),
\end{equation}
where
$\lambda(\x,\bu_d)=K^{-1}_y G_d k(\bu_d, \bu) $ with data covariance
$K_y=\mbox{Cov}(\bY_d,\bY_d)=G_d K(\bu_d,\bu_d)G_d^T+R_d$ and cross-covariance $\mbox{Cov}(\bY_d,Z(\x))=G_d k(\bu_d, \x)$, assuming the measurement
error is independent of the random fields.  The Kriging (cross-)covariance function can also be expressed in the same vein via
a $p \times p$ matrix 
\begin{equation}\label{eq:krig:cov}
\Sigma(\x,\x')=K(\x,\x')-\lambda(\x,\bu_d)^T K_y(\bu_d, \bu_d) \lambda(\x,\bu_d).
\end{equation}

@@

A critical element in the derivation of a closed-form for the EIBV is
that the conditional mean in Eq. \eqref{eq:cokrig} is a linear
(affine) function of the data $\by_d$ and the covariance is not a
function of the data. This in turn means that the probabilities in Eq.
\eqref{eq:post_ep} involve inequality statements for linear
combinations of Gaussian variables. Related closed-form solutions have
been noted in similar contexts \citep{bhattacharjya2013value,
  chevalier2014fast,stroh}, but not generalized to our situation with
random sets for vector-valued GPs. Notably, the entry in Eq.
\eqref{eq:post_ep} is in the form of $\ba+B V$, where
$V \sim N(0_q, K_y (\bu_d,\bu_d))$,
$\ba=(t_1,t_2)^t-\bmu(\x)$
and $B=\lambda(\x,\bu_d)^T$. The following proposition then applies to give a
closed-form solution to EIBV calculation in this case.

@@

The design and data define
the conditional GP with mean
$E(\gp[\x]|\mathcal{Y}^{s-1})$ and (cross) covariance matrix
$\mbox{Var}(\gp[\x]|\mathcal{Y}^{s-1})$. These are efficiently updated in a sequential manner keeping track of the mean and covariance kernel of the GP over the stages. As extensions of Eq. \eqref{eq:cokrig} and \eqref{eq:krig:cov}, we have
\begin{eqnarray}\label{eq:krigSUR}
\mu^{s}(\x) &=&\mu^{s-1}(\x)+\lambda^{[s,s]}(\x,\bu_{d^{j,s}})^T (\by_{d^{j,s}}-G_{d^{j,s}} \mu(\x)), \nonumber \\
k^{s}(\x,\x') &=& k^{s-1}(\x,\x')-\lambda^{[s,s]}(\x,\bu_{d^{j,s}})^T K_{Y_{d^{j,s}}} \lambda^{[s,s]}(\x',\bu_{d^{j,s}}),
\end{eqnarray}
where $\lambda^{[s,s]}(\x,\bu_{d^{j,s}})=K^{-1}_{Y_{d^{j,s}}} G_{d^{j,s}} k(\bu_{d^{j,s}},\x)$ and the data covariance is $K^{-1}_{Y_{d^{j,s}}}=G_{d^{j,s}} K^s(\bu_{d^{j,s}},\bu_{d^{j,s}})G^T_{d^{j,s}}+R_{d^{j,s}}$, with measurement model as in Eq. \eqref{eq:meas}.

At stage $s$, the robot must then choose design $d^{j,s}$ where measurements $\by_{d^{j,s}}$ should
be gathered. For the triangularized waypoint
graph all six possible
directions must be evaluated to minimize the expected uncertainty of random sets, and in particular the EIBV for our 
application. We hence focus on choosing designs for data modeled by
Eq. \eqref{eq:meas}, to conduct efficient sequential uncertainty
reduction according to the EIBV criterion. Compared with the
static design $d$ in Eq. \eqref{eq:eibv}, the optimal sequential
design not only considers the best current design grid point, but also
how the data gathered at this node could inform future sampling at
successive stages.

@@

Even though the strategy looks at two sampling stages, it is only used to find
the current best design. When data are collected at stage $s$, the GP model is
updated as in Eq. \eqref{eq:krigSUR}, and the mean and (cross) covariance are
used to compute the next design for stage $s+1$, now anticipating what
stage $s+1$ and $s+2$ could offer, and so on.
This look-ahead approach is much more computationally demanding than
the myopic strategy, so for practical implementation onboard the AUV this is reaching the limits allowed for calculation. We conduct some pruning of the least valuable design paths in the evaluation of Eq. \eqref{critLA}, to get a satisfactory computation time below one minute per waypoint. This means that we use the myopic
strategy to rank the three best designs following the first stage alone, and
for each of these preferred designs, we undertake the look-ahead calculations.
